---
title: "New York Open Statistical Meetup"
author: "Hernando Cortina"
date: "hch@alum.mit.edu"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(hs_write_cache = TRUE)
```

## Abstract
# Hockeystick: Visualizing Global Heating with R

2023 is on track to be the hottest year in human history. Public interest in monitoring and tracking climate data is increasing rapidly. In this talk we will review the `hockeystick` R package designed to easily import and visualize climate data. There are a multitude of publicly-available climate data sources from a large variety of academic and scientific organizations around the world. However, for casual users without knowledge of the sources navigating these datasets can be daunting and error-prone.  

The goal of the `hockeystick`^[hockeystick: Download and Visualize Essential Climate Change Data. https://cran.r-project.org/package=hockeystick; https://cortinah.github.io/hockeystick/] R package is to make essential global heating datasets easily available to non-climate experts. `hockeystick` users can download the latest raw data from authoritative sources as well as view it via pre-defined ggplot2 charts. Datasets include atmospheric CO~2~ and CH~4~, carbon emissions, instrumental, reconstructed, and paleo ice-core temperature records, sea levels, hurricanes, and Arctic/Antarctic sea-ice. Daily land and oceans temperatures were recently added. All data is compiled from authoritative sources, traceable, and documented.  `hockeystick` does not store any data itself, but serves as an API wrapper to always obtain the most up-to-date data. Users may work with the data frames returned by the package or use the pre-defined visualizations.

During the talk I will provide a short overview of the history of climate science and demonstrate this CRAN package, including a live demo. 

## Package demo

Retrieve NOAA/ESRL Mauna Loa CO~2~ Observatory concentration data and plot:

```{r atlantic, fig.retina=2, message=FALSE, out.width='75%'}
library(hockeystick)
atlantictemp <- get_dailytemp(region='AS')
plot_dailytemp(atlantictemp, anomaly = F)
```


Retrieve NOAA/ESRL Mauna Loa CO~2~ Observatory concentration data and plot:

```{r carbon, fig.retina=2, message=FALSE, out.width='75%'}
ml_co2 <- get_carbon()
plot_carbon(ml_co2)
```

Retrieve GCP global CO~2~ emissions and plot:

```{r emissions, out.width='75%', fig.retina=2, message=FALSE}
emissions <- get_emissions()
plot_emissions(emissions)
```

\newpage
  
```{r emissions2, out.width='75%', fig.retina=2, message=FALSE}
plot_emissions_with_land(emissions)
```


Visualize cumulative emissions by country:
```{r emissionsmap, out.width='75%', fig.retina=2}
emissions_map()
```

\newpage

Retrieve NASA/GISS global surface temperature anomaly data and plot:

```{r temp, out.width='75%', fig.retina=2}
anomaly <- get_temp()
plot_temp(anomaly)
```

Visualize warming using Ed Hawkins styled "warming stripes":
```{r stripes, out.width='75%', fig.retina=2}
warming_stripes()
```

\newpage 

Retrieve tide gauge and satellite sea level data and plot:

```{r sl, out.width='75%', fig.retina=2}
gmsl <- get_sealevel()
plot_sealevel(gmsl)
```

Retrieve July annual Arctic Sea Ice Index and plot:

```{r si, fig.retina=2, message=FALSE, out.width='75%', warning=FALSE}
seaice <- get_seaice()
plot_seaice(seaice)
```

`get_seaice()` arguments can be modified to download Antarctic sea ice, and allow any month.

\newpage

You can also visualize sea ice by month and year:

```{r icecurves, fig.retina=2, message=FALSE, out.width='75%'}
arcticice <- get_icecurves()
plot_icecurves(arcticice)
```

Retrieve Common Era temperature reconstruction and plot it with instrumental record:

```{r 2ktemp, out.width='75%', fig.retina=2}
anomaly2k <- get_temp2k()
plot_temp2k(anomaly2k)
```

\newpage

Retrieve NOAA HURDAT2 hurricane data and plot:

```{r hurricanes, out.width='75%', fig.retina=2, message=FALSE}
hurricanes <- get_hurricanes()
plot_hurricanes(hurricanes)
```

```{r hurricanes2, out.width='75%', fig.retina=2, message=FALSE}
plot_hurricane_nrg(hurricanes)
```

\newpage

Retrieve NOAA/ESRL CH~4~ Globally averaged mean data and plot:

```{r methane, fig.retina=2, message=FALSE, out.width='75%'}
ch4 <- get_methane()
plot_methane(ch4)
```

Retrieve Vostok paleo ice core data and plot:

```{r paleo, out.width='75%', fig.retina=2, message=FALSE}
vostok <- get_paleo()
plot_paleo(vostok)
```


## Acknowledgments
- Carbon Dioxide concentrations: Dr. Pieter Tans, NOAA/GML (https://gml.noaa.gov/ccgg/trends/) and Dr. Ralph Keeling, [Scripps Institution of Oceanography](https://scrippsco2.ucsd.edu/).
- Global temperature anomaly: *GISS Surface Temperature Analysis (GISTEMP), version 4.* GISTEMP Team, 2020: NASA Goddard Institute for Space Studies. https://data.giss.nasa.gov/gistemp/
- Warming Stripes design from Ed Hawkins' Climate Lab. https://www.climate-lab-book.ac.uk/2018/warming-stripes/. In addition grateful to Dr. Dominic Royé for posting his approach to plotting them using ggplot2, which `warming_stripes()` is based on. https://dominicroye.github.io/en/2018/how-to-create-warming-stripes-in-r/
- Sea level data: [NOAA Laboratory for Satellite Altimetry (sat)](https://www.star.nesdis.noaa.gov/socd/lsa/SeaLevelRise/) and [Commonwealth Scientific and Industrial Research Organisation (tide gauges)](https://research.csiro.au/slrwavescoast/sea-level/measurements-and-data/sea-level-data/)
- Sea Ice Index: [National Snow & Ice Data Center](https://nsidc.org/home).
Data Archive: https://nsidc.org/data/explore-data
- Vostok carbon dioxide and temperature data: https://cdiac.ess-dive.lbl.gov/trends/co2/vostok.html
- Common Era reconstructed temperature data: [PAGES2k Consortium and NOAA](https://www.ncei.noaa.gov/access/paleo-search/study/26872)). 
- Hurricanes: National Oceanic and Atmospheric Administration HURDAT Atlantic Hurricane Database [Re-analysis Project](https://www.aoml.noaa.gov/hrd/hurdat/Data_Storm.html), particularly Dr. Chris Landsea.
- Carbon Dioxide emissions: [Global Carbon Project](https://www.globalcarbonproject.org/carbonbudget/) and [Our World In Data](https://github.com/owid/co2-data)
- Methane: Lan, X., K.W. Thoning, and E.J. Dlugokencky, NOAA Global Monitoring Laboratory (https://gml.noaa.gov/ccgg/trends_ch4/).
- Thank you to Dirk Eddelbuettel for providing the .isConnected function from his [tint](https://CRAN.R-project.org/package=tint) package to test for internet connectivity.


